---
title: "GDM Prediction: Data Exploration and Pre-processing"
author: "Arne Nyeck Nyeck & Katie Schuman"
date: "2025-06-10"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Document Summary
### 1. Data Exploration
### 2. Data Cleaning
### 3. Data Pre-processing
### 4. Feature Selection


# Data Exploration

```{r}
library(ggplot2)
library(dplyr)
library(corrplot)
library(reshape2)
library(RColorBrewer)
install.packages(c("ggplot2", "reshape2"))
```

### Read Data File
```{r}
patients <- read.csv("https://raw.githubusercontent.com/arnenyeck06/Statistical-Inference-Package/refs/heads/main/patients.csv")
```

### Examining data structure
```{r}
head(patients)
summary(patients)
colnames(patients)
str(patients)
```

### Check missing values 
```{r}
colSums(is.na(patients))
```
### Examining positively diagnosed records
```{r}
Positive <- subset(patients, Diagnosis == 1)
head(Positive, 5)
```

### Visualizing Diagnosis
```{r}
ggplot(patients, aes(x = factor(Diagnosis))) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Count of Gestational Diabetes Diagnoses",
    x = "Diagnosis (0 = No, 1 = Yes)",
    y = "Count"
  ) +
  theme_minimal()
```


### Checking for skewness

### Blood Pressure
```{r}
ggplot(patients, aes(x = BloodPressure, fill = ..count..)) +
  geom_histogram(binwidth = 10, color = "black") +
  scale_fill_viridis_c() +
  labs(title = "Histogram of Blood Pressure",
       x = "Blood Pressure",
       y = "Count") +
  theme_minimal()
```

We can say that blood pressure is skewed left by looking at the graph.

### Skin Thickness
```{r}
ggplot(patients, aes(x = SkinThickness, fill = ..count..)) +
  geom_histogram(binwidth = 10, color = "black") +
  scale_fill_viridis_c() +
  labs(title = "Skin Thickness Histogram",
       x = "Skin Thickness",
       y = "Count") +
  theme_minimal()
```

We can say that skin thickness is skewed left by looking at the graph.

### Insulin
```{r}
ggplot(patients, aes(x = Insulin, fill = ..count..)) +
  geom_histogram(binwidth = 10, color = "black") +
  scale_fill_viridis_c() +
  labs(title = "Insulin Histogram",
       x = "Insulin",
       y = "Count") +
  theme_minimal()
```

We can say that Insulin is skewed left by looking at the graph.


### BMI
```{r}
ggplot(patients, aes(x = BMI, fill = ..count..)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_fill_viridis_c() +
  labs(title = "Histogram of BMI",
       x = "BMI",
       y = "Count") +
  theme_minimal()
```

There is an observed concentration between 20-40


### Diagnosis and Glucose
```{r}
ggplot(patients, aes(x = factor(Diagnosis), y = Glucose, fill = factor(Diagnosis))) +
  geom_boxplot(color = "black") +
  scale_fill_manual(values = c("0" = "#56B4E9", "1" = "#E69F00"),
                    labels = c("No GDM", "GDM")) +
  labs(
    title = "Glucose Levels by Diagnosis",
    x = "Gestational Diabetes Diagnosis",
    y = "Glucose (mg/dL)",
    fill = "Diagnosis"
  ) +
  theme_minimal()
```
Patients diagnosed with GDM have higher Glucose levels. 


### Plot variables with each other

### Blood pressure and glucose
```{r}
ggplot(Positive, aes(x = BloodPressure, y = Glucose)) +
  geom_point(color = "blue") +
  labs(title = "BloodPressure & Glucose",
       x = "BloodPressure",
       y = "Glucose") +
  theme_minimal()
```

### Age and Glucose
```{r}
ggplot(patients, aes(x = Age, y = Glucose, color = as.factor(Diagnosis))) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     name = "Diagnosis",
                     labels = c("Negative", "Positive")) +
  labs(title = "Glucose vs Age Colored by Diagnosis",
       x = "Age",
       y = "Glucose") +
  theme_minimal()
```

### BloodPressure and Age
```{r}
ggplot(patients, aes(x = Age, y = BloodPressure, color = as.factor(Diagnosis))) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     name = "Diagnosis",
                     labels = c("Negative", "Positive")) +
  labs(title = "BloodPressure vs Age Colored by Diagnosis",
       x = "Age",
       y = "BloodPressure") +
  theme_minimal()
```

### SkinThickness and Age
```{r}
ggplot(patients, aes(x = Age, y = SkinThickness, color = as.factor(Diagnosis))) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     name = "Diagnosis",
                     labels = c("Negative", "Positive")) +
  labs(title = "SkinThickness vs Age Colored by Diagnosis",
       x = "Age",
       y = "SkinThickness") +
  theme_minimal()
```

### Insulin and Age
```{r}
ggplot(patients, aes(x = Age, y = Insulin, color = as.factor(Diagnosis))) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     name = "Diagnosis",
                     labels = c("Negative", "Positive")) +
  labs(title = "Insulin vs Age Colored by Diagnosis",
       x = "Age",
       y = "Insulin") +
  theme_minimal()
```

### BMI and Age
```{r}
ggplot(patients, aes(x = Age, y = BMI, color = as.factor(Diagnosis))) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     name = "Diagnosis",
                     labels = c("Negative", "Positive")) +
  labs(title = "BMI vs Age Colored by Diagnosis",
       x = "Age",
       y = "BMI") +
  theme_minimal()
```

### Checking correlation between features
```{r}
# Assuming your data frame is called 'patients'
cor_matrix <- cor(patients[, sapply(patients, is.numeric)], use = "pairwise.complete.obs")
print(cor_matrix)


# Starting from patients data frame:
patients_corr <- patients[, sapply(patients, is.numeric)]

# Now calculate correlation matrix
cor_mat <- cor(patients_corr, use = "pairwise.complete.obs")

corrplot(cor_mat, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45)

```

### let's add some values
```{r}
# Now use brewer.pal
cols <- colorRampPalette(brewer.pal(9, "YlGnBu"))(100)

# Then plot
corrplot::corrplot(cor_matrix, method = "color", type = "upper",
                   tl.cex = 0.8,
                   addCoef.col = "black",
                   col = cols)
```



# Data Cleaning

Reading data
```{r}
library(readr)
datap <- read_csv("C:/Users/katie/Documents/CIS-690/patients.csv")
summary(datap)
str(datap)
```

### Replace 0s with NA in selected columns (where 0 indicates missing)
```{r}
# Leaving pregnancies zeros and Diagnosis zeros as these are meaningful
datap$Insulin[datap$Insulin == 0] <- NA
datap$BMI[datap$BMI == 0] <- NA
datap$SkinThickness[datap$SkinThickness == 0] <- NA
datap$BloodPressure[datap$BloodPressure == 0] <- NA
datap$Pedigree[datap$Pedigree == 0] <- NA
datap$Glucose[datap$Glucose == 0] <- NA
datap$Age[datap$Age == 0] <- NA
```

### Explore NA values
```{r}
#Count NAs
sum(is.na(datap))
#652 NAs total
sum(apply(datap, 1, function(row) any(is.na(row))))
#376 rows have a missing value
sum(datap == 0, na.rm = TRUE)
#611 zeros in total 
```

### Double checking that conversion was successful for each column
```{r}
sum(datap$Age == 0, na.rm = TRUE)
#All Age zeros converted to NA
sum(datap$SkinThickness == 0, na.rm = TRUE)
#All SkinTckness zeros converted to NA
sum(datap$Insulin == 0, na.rm = TRUE)
#All Insulin zeros have been converted to NA
sum(datap$BMI == 0, na.rm = TRUE)
#All BMI zeros have been converted to NA
sum(datap$Pedigree == 0, na.rm = TRUE)
#All pedigree zeros have been converted to NA
sum(datap$Glucose == 0, na.rm = TRUE)
#All glucose zeros have been converted to NA
sum(datap$BloodPressure == 0, na.rm = TRUE)
#All blood pressure zeros have been converted to NA
sum(datap$Diagnosis == 0, na.rm = TRUE)
sum(datap$Diagnosis == 1, na.rm = TRUE)

```

Diagnosis and pregnancy zeros have been left as zeros (Binary- positive or negative diagnosis).

268 postive diagnosis in the dataset
500 negative diagnosis in the dataset

### Checking Correlation
```{r}
#correlation matrix
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("corrplot")
library(corrplot)

#Ignore NAs and visualize correlation between variables
cor_matrix2 <- cor(datap[, 1:9], use = "complete.obs")
print(cor_matrix2)

corrplot(cor_matrix2, method = "circle")
#Highly correlated features: 
#Age & number of pregnancies- 0.68
#Glucose & Insulin- 0.58
#BMI & skin thickness- 0.66
#Diagnosis & glucose- 0.515

install.packages("car")
library(car)

#VIF
model <- lm(Diagnosis ~ BMI + Age + Pregnancies + Pedigree + Glucose + BloodPressure, Insulin, SkinThickness, data = datap)

vif(model)

```
There are some correlated values, but VIF scores are all below 3. 

### Investigate missingness
```{r}
install.packages("naniar")
library(naniar)
# Show rows with any missing data
datap[!complete.cases(datap), ]
# Visualize missingness by variable
gg_miss_var(datap)
# Visualize missing data patterns
vis_miss(datap)
```

### NA & Zeros Totals
```{r}
colSums(is.na(datap))
colSums(datap == 0, na.rm = TRUE)
#3 columns with <35 missing rows
#111 rows where pregnancies == 0
#500 rows where Diagnosis == 0
```


### Find rows where this count is 3 or more
```{r}
zero_or_na_per_row <- apply(datap, 1, function(row) {
  sum(is.na(row))
})

(rows_3_or_more <- sum(zero_or_na_per_row >= 3))
```
We see that there are 35 rows where blood pressure, skin thickness, and insulin are missing together. We choose to remove these rows at less than 4% of the data. 

### Remove rows with 3 or more NAs
```{r}
# Remove rows with 3 or more NAs
df_clean <- datap[rowSums(is.na(datap)) < 3, ]
```


### Vizualize missingness together
```{r}
gg_miss_upset(datap)
```

SkinThickness and Insulin missing together 192 times. 
Possible missingness explainations: more invasive testing, or these items are only done for certain clients or appointment types. This should be explored to see if data is MAR, MCAR, or MNAR.

### Check if both missing is related to other variables
1. T-tests between each variable and missingness
2. Visualize to observe any differences


### 1. 
```{r}
length(datap$Insulin)
length(datap$SkinThickness)

datap$insulin_skin_missing <- ifelse(is.na(datap$Insulin) & is.na(datap$SkinThickness), 1, 0)

# Run t-test for BMI
t_result <- t.test(BMI ~ insulin_skin_missing, data = datap)
print(t_result)

# Run t-test for Pregnancies
t_result <- t.test(Pregnancies ~ insulin_skin_missing, data = datap)
print(t_result)

# Run t-test for Age
t_result <- t.test(Age ~ insulin_skin_missing, data = datap)
print(t_result)

# Run t-test for Pedigree
t_result <- t.test(Pedigree ~ insulin_skin_missing, data = datap)
print(t_result)

# Run t-test for Glucose
t_result <- t.test(Glucose ~ insulin_skin_missing, data = datap)
print(t_result)

# Run t-test for BloodPressure
t_result <- t.test(BloodPressure ~ insulin_skin_missing, data = datap)
print(t_result)
```

### 2. 
```{r}

library(ggplot2)

ggplot(datap, aes(x = factor(insulin_skin_missing), y = Age)) +
  geom_boxplot(fill = c("#a6cee3", "#1f78b4")) +
  labs(x = "Insulin & SkinThickness Missing (0 = No, 1 = Yes)", y = "Age",
       title = "Age Distribution by Missingness of Insulin and SkinThickness") +
  theme_minimal()

ggplot(datap, aes(x = factor(insulin_skin_missing), y = Pregnancies)) +
  geom_boxplot(fill = c("#fdbf6f", "#ff7f00")) +
  labs(x = "Insulin & SkinThickness Missing (0 = No, 1 = Yes)", y = "Number of Pregnancies",
       title = "Pregnancies Distribution by Missingness of Insulin and SkinThickness") +
  theme_minimal()

ggplot(datap, aes(x = factor(insulin_skin_missing), y = Pedigree)) +
  geom_boxplot(fill = c("blue", "forestgreen")) +
  labs(x = "Insulin & SkinThickness Missing (0 = No, 1 = Yes)", y = "Pedigree",
       title = "Pedigree Distribution by Missingness of Insulin and SkinThickness") +
  theme_minimal()
```


Analysis: There is a statistically significantly difference in several features means and missing or not missing the set of variables. We are suspecting that high missing rates of skin thickness and insulin are missing at random. Upon researching the domain, we suspect that older patients, or patients who have been pregnant a number of times before may often  skip these measurements. They are not routine, and are possibly opted for only when there are other predictive factors observed. This indicates the missingness is at random (MAR). Imputation methods should be carefully considered using this information. 


### Are zeros in 'Pregnancies' errors? Or does this column ask about previous pregnancies, not including the current pregnancy?
```{r}
# Subset rows where Pregnancies == 0
preg0_df <- subset(datap, Pregnancies == 0)

# Count how many of those have Diagnosis == 1 (GDM)
gdm_with_preg0 <- sum(preg0_df$Diagnosis == 1, na.rm = TRUE)

# Total number of rows with Pregnancies == 0
total_preg0 <- nrow(preg0_df)

# Print results
cat("Total rows with Pregnancies == 0:", total_preg0, "\n")
cat("Rows with Pregnancies == 0 AND Diagnosis == 1 (GDM):", gdm_with_preg0, "\n")
```
There are 38 cases where pregnancy ==0 and diagnosis ==1. 
We will assume 'Pregnancies' == 0 means prior pregnancies. 



### Check for duplicates
```{r}
any(duplicated(df_clean))
```

### Recheck NA's
```{r}
colSums(is.na(df_clean))
```


### Identify and remove columns with fewer than 6 NAs
```{r}
cols_to_clean <- names(which(colSums(is.na(df_clean)) < 6))
cols_to_clean

# Keep rows with no NA in those selected columns
df_clean <- df_clean[complete.cases(df_clean[, cols_to_clean]), ]
```

### Recheck
```{r}
colSums(is.na(df_clean))
```
The only columns with NAs remaining are SkinThickness and Insulin

### Reviewing rows remaining and remaining missingness
```{r}
#Total number of rows
(total_rows <- nrow(df_clean))
#729 rows remaining after basic cleaning
# Percentage of missing SkinThickness
skin_missing_pct <- sum(is.na(df_clean$SkinThickness)) / total_rows * 100
# Percentage of missing Insulin
insulin_missing_pct <- sum(is.na(df_clean$Insulin)) / total_rows * 100
# Display the results
cat("Percentage of missing SkinThickness values:", round(skin_missing_pct, 2), "%\n")
cat("Percentage of missing Insulin values:", round(insulin_missing_pct, 2), "%\n")
```
27% of Skin thickness values are missing.
46% of Insulin values are missing. 

### Checking for outliers
```{r}
vars_to_plot <- c("Insulin", "BMI", "SkinThickness", "BloodPressure", "Pedigree", "Glucose", "Age")

par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))

for (var_name in vars_to_plot) {
  boxplot(df_clean[[var_name]], 
          main = paste(var_name, "Boxplot"),
          horizontal = TRUE, 
          col = "lightblue",
          na.action = na.omit)
}
```

### Count and examine problematic values
```{r}
# BloodPressure issues (0 or < 40)
bp_issues <- df_clean[ df_clean$BloodPressure < 40, ]
cat("BloodPressure < 40:", sum(df_clean$BloodPressure < 40, na.rm = TRUE), "\n")
cat("Total BloodPressure issues:", nrow(bp_issues), "\n\n")
```
To remove 4 instances of blood pressure errors

```{r}
# SkinThickness issues (0 values)
skin_issues <- df_clean[df_clean$SkinThickness < 5 | df_clean$SkinThickness > 60, ]
cat("SkinThickness < 5:", sum(df_clean$SkinThickness < 5, na.rm = TRUE), "\n")
cat("SkinThickness > 60:", sum(df_clean$SkinThickness > 60, na.rm = TRUE), "\n")
```
To remove 2 Instances of skin thickness errors

```{r}
# Glucose issues (< 50)
glucose_issues <- df_clean[df_clean$Glucose < 50, ]
cat("Glucose < 50:", nrow(glucose_issues), "\n\n")
```
To remove 6 instances of glucose errors

```{r}
#Age
print(which(df_clean$Age > 50))
print(which(df_clean$Age < 11))
#Inspect rows
print(as.data.frame(df_clean[c(8,  11,  12,  13,  21,  25,  27,  36,  40,  49,  61,  84, 106, 114, 119, 130, 137, 173,
                   190, 192, 193, 198, 207, 208, 221, 244, 247, 256, 260, 266, 267, 276, 281, 300, 322, 336,
                   337, 338, 349, 362, 375, 426, 428, 429, 444, 448, 455, 457, 462, 465, 475, 478, 482, 483,
                   485, 500, 505, 509, 515, 520, 523, 544, 546, 550, 561, 613, 615, 621, 629, 631, 655, 668,
                   670, 685, 708, 710, 714), ]))
                   
# Specify the row indices to remove
rows_to_remove <- c(8, 11, 12, 13, 21, 25, 27, 36, 40, 49, 61, 84, 106, 114, 119, 130, 137, 173,
                    190, 192, 193, 198, 207, 208, 221, 244, 247, 256, 260, 266, 267, 276, 281, 300,
                    322, 336, 337, 338, 349, 362, 375, 426, 428, 429, 444, 448, 455, 457, 462, 465,
                    475, 478, 482, 483, 485, 500, 505, 509, 515, 520, 523, 544, 546, 550, 561, 613,
                    615, 621, 629, 631, 655, 668, 670, 685, 708, 710, 714)

# Create new cleaned dataset by removing those rows
#df_clean_2 <- df_clean[-rows_to_remove, ]

print(df_clean[c(419, 545, 16,  58, 118, 562, 564), ])
#row 419 looks normal comparing to bmi- keep

# Remove rows 545, 16, 58, 118, 562, 564
#rows_to_remove <- c(545, 16, 58, 118, 562, 564)

# Create new dataframe without these rows
#df_clean_2 <- df_clean_2[-rows_to_remove, ]

```


### Concise removal of non-realistic values
```{r}
library(dplyr)
df_clean_2 <- df_clean %>%
  filter(BloodPressure >= 40) %>%
  filter(SkinThickness >= 5 & SkinThickness <= 60) %>%
  filter(Glucose >= 50) %>%
  filter(Age >= 11 & Age <= 50)
```

```{r}
library(dplyr)
df_clean_2 <- df_clean %>%
  dplyr::filter(Age >= 11 & Age <= 50,
                Glucose >= 50,
                SkinThickness >= 5 & SkinThickness <= 60,
                BloodPressure >= 40)
```


### Check
```{r}
summary(df_clean_2)
str(df_clean_2)
```


# Imputation

### Create two different sets of data: One with Insulin and SkinThickness imputed with RF, and the other imputed with MICE

### Mice Imputation
```{r, error=TRUE}
install.packages("mice")
library(mice)
data_to_be_imputed = df_clean_2
imputed_data <- mice(data_to_be_imputed, method = "pmm", m = 5, seed = 123)
imputed_df <- complete(imputed_data)  # gets the first completed dataset

# Check NAs
colSums(is.na(imputed_df))
summary(imputed_df)
head(imputed_df)

#Final imputed dataset
rf_dataset_imputed <- imputed_df

write.csv(rf_dataset_imputed, "completed_data_mice.csv", row.names = FALSE)
```


### RF Imputation
```{r, error=TRUE}
install.packages("missForest")
library(missForest)

set.seed(123)  # For reproducibility

# Save outcome variable (Diagnosis) separately
outcome_var <- data_to_be_imputed$Diagnosis

# Remove outcome before imputation
df_features_only <- data_to_be_imputed[, !(names(data_to_be_imputed) %in% "Diagnosis")]

# Check Insulin and SkinThickness presence before imputation
stopifnot("Insulin" %in% colnames(df_features_only))
stopifnot("SkinThickness" %in% colnames(df_features_only))

# Perform RF imputation on features only
rf_imputed <- missForest(df_features_only, maxiter = 10, ntree = 100)

# Extract completed data
rf_dataset_imputed <- rf_imputed$ximp

# Add outcome variable back
rf_dataset_imputed$Diagnosis <- outcome_var

# Check imputation results
colSums(is.na(rf_dataset_imputed))
summary(rf_dataset_imputed)
head(rf_dataset_imputed)

write.csv(rf_dataset_imputed, "completed_data_rf.csv", row.names = FALSE)
```


### Imputed Datasets- Github
```{r}
completed_data_mice <- read.csv("https://raw.githubusercontent.com/arnenyeck06/Statistical-Inference-Package/refs/heads/main/completed_data_mice.csv")

completed_data_rf <-
read.csv("https://raw.githubusercontent.com/arnenyeck06/Statistical-Inference-Package/refs/heads/main/completed_data_rf.csv")
```



# Pre-Processing

# Scaled dataset

SVM can benefit from feature transformation of highly skewed features. A data set will be created using log transformation on several of the features. 

### Normality Test
```{r}
rfid <- completed_data_rf
mid <- completed_data_mice

shapiro.test(rfid$Pedigree)
#not normal
shapiro.test(rfid$Age)
#not normal
shapiro.test(rfid$Insulin)
#not normal
shapiro.test(rfid$Glucose)
#not normal
shapiro.test(rfid$BMI)
#not normal
shapiro.test(rfid$BloodPressure)
#not normal
shapiro.test(rfid$SkinThickness)
#not normal
shapiro.test(rfid$Pregnancies)
#not normal


install.packages("e1071")

# Load
library(e1071)

# Check skewness for each numeric column
sapply(rfid, skewness, na.rm = TRUE)

#RF dataset transformations
rfid_log_transformed <- rfid %>%
  mutate(
    log_Insulin = log(Insulin + 1),        
    log_Pedigree = log(Pedigree + 1)
  )

#MICE dataset transformations
mid_log_transformed <- mid %>%
  mutate(
    log_Insulin = log(Insulin + 1),        
    log_Pedigree = log(Pedigree + 1)
  )
```

Log transformations were applied on the two features with highest skewness, Pedigree and Insulin. 



# Feature Selection

Models to be run based on the literature review are XGBoost, Logistic regression, Random Forest, and SVM. XGBoost and Random Forest handle feature selection internally. Logistic regression and SVM can benefit from feature selection. SVM feature selection will be handled here using Recursive Feature Elimination (RFE). This will be performed using both styles of imputation to compare performance between them. 

```{r}
# Read dataset
## ALL FOUR
library(caret)
set.seed(123)

data1 <- rfid_log_transformed
data2 <- mid_log_transformed


#Predictors for each dataset
x_rf <- data1[, c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
                 "BMI", "Age", "log_Pedigree", "log_Insulin")]
y_rf <- factor(data1$Diagnosis, levels = c(0, 1), labels = c("No", "Yes"))

x_mice <- data2[, c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
                    "BMI", "Age", "log_Pedigree", "log_Insulin")]
y_mice <- factor(data2$Diagnosis, levels = c(0, 1), labels = c("No", "Yes"))

#CV
train_ctrl <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  sampling = "smote"
)

rfe_ctrl <- rfeControl(functions = caretFuncs, method = "cv", number = 3)


#RFE SVM
rfe_svm_rf <- rfe(
  x = x_rf, y = y_rf,
  sizes = 1:8,
  rfeControl = rfe_ctrl,
  method = "svmRadial",
  metric = "ROC",
  trControl = train_ctrl
)

rfe_svm_mice <- rfe(
  x = x_mice, y = y_mice,
  sizes = 1:8,
  rfeControl = rfe_ctrl,
  method = "svmRadial",
  metric = "ROC",
  trControl = train_ctrl
)

# Create data frames of results for plotting
get_perf_df <- function(rfe_obj, model_name) {
  data.frame(
    Variables = rfe_obj$results$Variables,
    Accuracy = rfe_obj$results$Accuracy,
    Model = model_name
  )
}

results_combined <- rbind(
  get_perf_df(rfe_svm_rf, "SVM (RF Imputed)"),
  get_perf_df(rfe_svm_mice, "SVM (MICE Imputed)")
)

# Plot accuracy vs. number of features
ggplot(results_combined, aes(x = Variables, y = Accuracy, color = Model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "RFE Accuracy Comparison",
       x = "Number of Selected Features",
       y = "ROC") +
  scale_color_brewer(palette = "Set1")

plot(rfe_svm_rf, type = c("g", "o"))
plot(rfe_svm_mice, type = c("g", "o"))
rfe_svm_rf$optVariables
rfe_svm_mice$optVariables
```

Rf returned as the highest performing dataset, with six features included. For our SVM model, we will run this subset of features with the random forest and compare with the full set of features. 

### Create a dataset with this subset of features & log transformation.
```{r}
#include only recommended features determined in feature selection:
rf_log_transf_feature_subset <- rfe_svm_rf[c("Glucose", "log_Insulin", "Age", "BMI", "SkinThickness", "Pregnancies", "Diagnosis")]

colnames(rf_log_transf_feature_subset)
#write.csv(rf_log_transf_feature_subset, "completed_data_rf_log_transformed_feature_subset.csv", row.names = FALSE)
```

### Dataset for SVM - Github
```{r}
completed_data_rf_log_transformed_feature_subset_for_SVM <- read.csv("https://raw.githubusercontent.com/arnenyeck06/Statistical-Inference-Package/refs/heads/main/completed_data_rf_log_transformed_feature_subset_for_SVM.csv")
```



